---
title: "Generate Figure 5"
author: "Thomas Veith"
date: "01/02/2021"
output: html_document
---

Set correct root.dir for the project:
(Note that code expects output of "Generate Figures 2-4.Rmd" inside root.dir. In addition surrogate mis-segregation rates for the nine cancer type must be under input/Predicted_LaggingChrPct_TCGA.csv)
```{r setup}
knitr::opts_knit$set(root.dir = "path/to/dir")
```

Load libraries and define constants:
```{r}
library(ggplot2) 
library(gridExtra) 
library(matlab)
library(magrittr)
N_SAMPLES=1000
WHICH_CURVE="constant"
ALPHA=0.35
SHADECOLOR="gray"
MAXCHROMO=32
```

Read predicted critical curves and plot one example:
```{r}
critcurves = read.csv(
  "expectedResults/all_critical_curves.csv",
  header = T,
  check.names = F,
  stringsAsFactors = F,
  strip.white = T
)
critcurves = critcurves[critcurves$maxchrom == MAXCHROMO,]
critcurves$departFromHomeostasis = 1 - critcurves$pop_avg_death_rate
critcurves$category = "Head and Neck"
new_critcurves = critcurves[critcurves$`B(i,beta)` == "B(i)==1/(beta+e^sqrt( abs( i-3 ) ))" &
                              critcurves$`D(i,mu)` == " D(i)==mu",]
old_critcurves = critcurves[critcurves$`B(i,beta)` == " B(i)==beta" &
                              critcurves$`D(i,mu)` == " D(i)==mu",]
tmp = old_critcurves[old_critcurves$pop_avg_death_rate < min(new_critcurves$pop_avg_death_rate),]
tmp$`B(i,beta)` = unique(new_critcurves$`B(i,beta)`)
tmp$pop_avg_misseg_rate = max(new_critcurves$pop_avg_misseg_rate)
new_critcurves = rbind(new_critcurves, tmp)
plot(
  new_critcurves$departFromHomeostasis,
  new_critcurves$pop_avg_misseg_rate,
  xlim = c(1E-3, 1),
  ylim = c(0, 1),
  log = "x",
  pch = 20
)
if (WHICH_CURVE == "constant") {
  critcurves = old_critcurves
} else{
  critcurves = new_critcurves
}
```

Read/format data needed to produce 2D box plots of Chromosomal Mis-segregation
1) Read log medians & sdNorms for birth & growth rates into data frame 'birth_and_growth_Rates'
2) Create 'gRLN', 'bRLN' lists of log norm distributed growth rates and birth rates per cancer type
3) Set up 'dOVERb' data frame of death/birth where death=1-growth/birth, and replace negative entries with 0
4) Read 'inp' lagging chromosome data, remove metastases data, and normalize
5) Create 'subMap' mapping diseases in 'inp$group' to 'birth_and_growth_Rates$type', and set former to match latter
6) Add 1-death/birth to 'dOVERb' data frame
7) Choose whether to plot death/birth or 1-death/birth
8) Set x and y-labels for box plots
```{r}

birth_and_growth_Rates <-
  data.frame(
    "blm" =
      c(
        -2.51405,
        -1.79009,
        -1.84878,
        -1.21294,
        -2.49839,
        -1.39956,
        -2.07355,
        -2.59514,
        -3.31508
      ),
    "bsd" =
      c(
        1.31437,
        0.655826,
        0.91377,
        0.194515,
        0.888437,
        0.162913,
        0.447807,
        0.991364,
        1.04925
      ),
    "glm" =
      c(
        -5.3441,
        -1.98498,
        -7.00215,
        -5.46545,
        -6.30794,
        -6.78071,
        -5.04616,
        -4.85863,
        -5.4622
      ),
    "gsd" =
      c(
        0.874367,
        0.477869,
        1.54467,
        0.578148,
        0.998794,
        1.06405,
        0.410637,
        0.246221,
        1.83844
      ),
    "type" =
      c(
        "Head and Neck",
        "Esophageal",
        "Colorectal",
        "Rectal",
        "Breast",
        "Cervix",
        "Skin (Melanoma)",
        "Lung (Small-Cell)",
        "Diffuse Large B-Cell"
      )
  )

gRLN <- list()
bRLN <- list()
for (i in 1:nrow(birth_and_growth_Rates)) {
  gRLN[[birth_and_growth_Rates$type[i]]] <-
    rlnorm(N_SAMPLES, meanlog = birth_and_growth_Rates$glm[i], sdlog =  birth_and_growth_Rates$gsd[i])
  bRLN[[birth_and_growth_Rates$type[i]]] <-
    rlnorm(N_SAMPLES, meanlog = birth_and_growth_Rates$blm[i], sdlog =  birth_and_growth_Rates$bsd[i])
}

dOVERb <- data.frame()
for (name in unique(birth_and_growth_Rates$type)) {
  x <- data.frame("db" = 1 - (gRLN[[name]] / bRLN[[name]]))
  x[x$db < 0,] <- 0
  x$type = repmat(name, N_SAMPLES, 1)
  dOVERb <- rbind(dOVERb, x)
}

inp <-
  read.table("input/Predicted_LaggingChrPct_TCGA.txt", header = T)
inp <- inp[inp$group != "met",]
inp$Lagging.Chromosome[inp$Lagging.Chromosome > 100] = 100
inp$Lagging.Chromosome <- inp$Lagging.Chromosome / 100

subMap = data.frame("inp" = sort(unique(inp$group)),
                    "type" = sort(unique(birth_and_growth_Rates$type)))
for (name in subMap$inp) {
  inp[inp$group == name, "group"] <-
    subMap[subMap$inp == name, "type"]
}

dOVERb$one_minus_d_over_b <- (1 - dOVERb$db)

## choose whether you want to plot 1-d/b(T) or d/b(F)
adjVal<-T
if (adjVal==T) {
  xlab = expression(paste("Departure from Homeostasis (1-",mu,"/",lambda, ")"))
} else{
  xlab = expression(paste(mu,"/",lambda))
} 
ylab = expression(paste("Mis-segregation Rate (", beta, ")"))
```

Generate two separate box plots (one for each dimension):
1) Create box plot for x dimension using death/birth or 1-death/birth
2) Create box plot for y dimension using lagging chromosome data
3) Produce 2-panel figure for visual verification of box plots
4) "Layer the data" (i.e., take what we want from the box plots to construct our geom_rect() objects below)
5) Create 'df' data frame that will be used by ggplot for the 2D box plot
```{r}
if (adjVal==T) {
  plot.x <-
    ggplot(dOVERb) + geom_boxplot(aes(type, one_minus_d_over_b))  + coord_flip()
} else {
  plot.x <-
    ggplot(dOVERb) + geom_boxplot(aes(type, db))  + coord_flip()
}
plot.y <-
  ggplot(inp) + geom_boxplot(aes(group, Lagging.Chromosome))  + coord_flip()
grid.arrange(plot.x, plot.y, ncol = 2) # visual verification of the box plots

plot.x <- layer_data(plot.x)[, 1:6]
plot.y <- layer_data(plot.y)[, 1:6]
colnames(plot.x) <- paste0("x.", gsub("y", "", colnames(plot.x)))
colnames(plot.y) <- paste0("y.", gsub("y", "", colnames(plot.y)))

df <- cbind(plot.x, plot.y)
df$category <- sort(unique(dOVERb$type))
```

Load both box plots into same plot:
```{r}
pdf(paste0("figs/05_",WHICH_CURVE,"-beta.pdf"), width = 7,height = 5)
ggplot(df, aes(fill = category, color = category)) +
  geom_line(
    data = critcurves,
    aes(x = departFromHomeostasis, y = pop_avg_misseg_rate),
    size = 1,
    color = "black",
    linetype = "dotted"
  )  +
  
  ## 2D box defined by the Q1 & Q3 values in each dimension, with outline
  geom_rect(aes(
    xmin = x.lower,
    xmax = x.upper,
    ymin = y.lower,
    ymax = y.upper
  ),
  alpha = 0.3) +
  geom_rect(
    aes(
      xmin = x.lower,
      xmax = x.upper,
      ymin = y.lower,
      ymax = y.upper
    ),
    color = "black",
    fill = NA
  ) +
  
  ## whiskers for x-axis dimension with ends
  geom_segment(aes(
    x = x.min,
    y = y.middle,
    xend = x.max,
    yend = y.middle
  )) + ## whiskers
  geom_segment(aes(
    x = x.min,
    y = y.lower,
    xend = x.min,
    yend = y.upper
  )) + ## lower end
  geom_segment(aes(
    x = x.max,
    y = y.lower,
    xend = x.max,
    yend = y.upper
  )) + ## upper end
  
  ## whiskers for y-axis dimension with ends
  geom_segment(aes(
    x = x.middle,
    y = y.min,
    xend = x.middle,
    yend = y.max
  )) + ## whiskers
  geom_segment(aes(
    x = x.lower,
    y = y.min,
    xend = x.upper,
    yend = y.min
  )) + ## lower end
  geom_segment(aes(
    x = x.lower,
    y = y.max,
    xend = x.upper,
    yend = y.max
  )) + ## upper end
  
  ylab(ylab) + xlab(xlab)+
  scale_x_continuous(trans = "log", breaks=c(0, 0.002, 0.05, 1))  +
  theme_classic() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(colour = "black", size = 4)
  ) +
  
  ## shaded areas
  geom_ribbon(data=critcurves,
              aes(x=departFromHomeostasis,y = pop_avg_misseg_rate,
                  ymin=pop_avg_misseg_rate,ymax=1),
              color = "gray",fill=SHADECOLOR,
              position="identity",alpha=ALPHA,linetype=2,) +
  
  ## shaded areas
  geom_ribbon(data=critcurves,
              aes(x=departFromHomeostasis,y = pop_avg_misseg_rate,
                  ymin=max(pop_avg_misseg_rate),ymax=1),
              color = "black",fill="black",
              position="identity",alpha=0.2,linetype=2)      

plot(0, type = 'n', axes = FALSE, ann = FALSE)
legend("center",c("Not a steady state","Steady state leads to extinction","Steady state leads to exponential growth"),fill=c("black","gray","white"),cex=1.5)
dev.off()
```
Acknowledgement to https://stackoverflow.com/questions/46068074/double-box-plots-in-ggplot2