---
title: "MIE heterogeneous mis-segr. rate"
author: "Richard J Beck"
date: "6/8/2022"
output: html_document
---

This file produces the figure related to predicting MIE when mis-segregation rates are heterogeneous - including showing the functional forms, plotting the critical curves, and evaluating the dynamics of the model

```{r setup}
knitr::opts_knit$set(root.dir = "PATH/TO/ROOT/DIR")
```

Source required functions and load libraries:
```{r,message=FALSE}
library(parallel)
library(deSolve)
library(pbapply)
library(R.utils)
library(ggplot2)
library(ggpubr)
library(stringr)
library(deSolve)
library(egg)
source("Rscripts/functions.R")
```

Generate data for 'linear' function
```{r}
## function to determine steady state mis-segregation rate
pop_ss_B <- function(beta,mu,maxchrom,B){
  A <- get_A(maxchrom=maxchrom,beta=beta,mu=mu,B=B)
  ss <- eigen(t(A))
  ss <- abs(ss$vectors[,apply(ss$vectors,2,function(i) prod(i>0)==1)])
  ss <- ss/sum(ss)
  p_mis <- sapply(1:length(ss), function(i) B(beta,i))
  sum(ss*p_mis)
}

## mis-segregation rate function
Blin <- function(beta,i){
  min(1,exp(beta-1.352857*i)/100)
  
}
## finds the lowest value of mis-segregation rate in the domain 
Bmin <- function(beta,B,minchrom,maxchrom){
  min(sapply(minchrom:maxchrom, function(i) B(beta,i)))
}

##produces data for the plot showing the behaviour of the function beta()
maxchrom <- 5
df_demo <- do.call(rbind,lapply(-2:4, function(beta){
  data.frame(theta=beta,ploidy=1:maxchrom,B=sapply(1:maxchrom, function(i) Blin(beta,i)))}))

## values of B to parameterise the function 
beta_vals <- -2.657704*seq(1,-2.25,-0.1625)

## find the critical curve
df_lin <- do.call(rbind,pblapply(beta_vals,find_c_crit,maxchrom=maxchrom,B=Blin))

## compute population avg mis-seg. rate
df_lin$pop_B <- sapply(1:nrow(df_lin), function(i){
  pop_ss_B(df_lin$beta[i],df_lin$mu[i],maxchrom=maxchrom,B=Blin)
})

## compute population minimum mis-seg. rate
df_lin$minB <- sapply(df_lin$beta, Bmin, B=Blin,minchrom=1,maxchrom=maxchrom)

## select parameters to run the ODE model
pars <- df_lin[nrow(df_lin)-2,]
beta <- pars$beta
mu <- pars$mu

ode_pars <- data.frame(beta=beta,mu=mu)

#run ode model
x <- run_ode(beta = beta,mu = mu,maxchrom=maxchrom,times=c(seq(0.1,10.9,0.1),11:99,seq(100,500,10)),founder=2,B=Blin)

# rescale to frequencies 
x[,2:ncol(x)]<-x[,2:ncol(x)]/rowSums(x[,2:ncol(x)])

## compile data for plotting
y <- reshape2::melt(x,id.vars="time")

lin_plot_data <- list(Bfunc_data = df_demo, c_crit_data = df_lin,ode_data=y ,ode_pars=ode_pars)
saveRDS(lin_plot_data,"data/lin_plot_data.Rds")

```
Generate data for sinusoidal function
```{r}
## function to determine steady state mis-segregation rate
pop_ss_B <- function(beta,mu,maxchrom,minchrom,B){
  A <- get_A(maxchrom=maxchrom,minchrom=minchrom,beta=beta,mu=mu,B=B)
  ss <- eigen(t(A))
  ss <- abs(ss$vectors[,apply(ss$vectors,2,function(i) prod(i>0)==1|prod(i<0)==1)])
  ss <- ss/sum(ss)
  p_mis <- sapply(1:length(ss), function(i) B(beta,minchrom+i-1))
  sum(ss*p_mis)
}
## mis-segregation rate function
Blog <- function(beta=1,i){
  b <- -0.09*i + beta*sin(45.59-12.25*log(i))
  min(1,exp(b)/100)
}

##produces data for the plot showing the behaviour of the function beta()
maxchrom <- 110
minchrom <- 22
df_demo <- do.call(rbind,lapply((4:9)/2, function(beta){
  data.frame(theta=beta,ploidy=minchrom:maxchrom,B=sapply(minchrom:maxchrom, function(i) Blog(beta,i)))}))



beta_vals <- 2.66*seq(0.7,1.7,0.025)


## find the critical curves but this takes long so we do in parallel so 
## set to FALSE to use pre-generated data
if(FALSE){
  cl <- makeCluster(getOption("cl.cores", 4))
  clusterCall(cl, function() { 
    source("Rscripts/functions.R")
    })
  df_log <- do.call(rbind,parLapplyLB(cl = cl, X=beta_vals, fun=fast_c_crit,
                                      maxchrom=maxchrom,minchrom=minchrom,B=Blog,df=TRUE))
  df_log$pop_B <- sapply(1:nrow(df_log), function(i){
  pop_ss_B(df_log$beta[i],df_log$mu[i],maxchrom=maxchrom,minchrom=minchrom,B=Blog)
})
  df_log$minB <- sapply(df_log$beta, Bmin, B=Blog,minchrom=minchrom,maxchrom=maxchrom)
  saveRDS(df_log,"data/log_ccrit.Rds")
}else{print("using saved data...")}

#load pre-saved (or just saved) data
df_log <- readRDS("data/log_ccrit.Rds")

##get pars for ode model
i <- 10
pars <- df_log[i,]
beta <- pars$beta
mu <- pars$mu
ode_pars <- data.frame(beta=beta,mu=mu)

# run ODE
#need the times to go up logarithmically otherwise the plots will look terrible,
# or the ODE model will never finish.
x <- run_ode(beta = beta,mu = mu,maxchrom=maxchrom,minchrom=minchrom,
             times=c(seq(100,1000,10),seq(1100,10000,100),
                     seq(11000,100000,1000),seq(110000,1000000,10000),
                     seq(1100000,10000000,100000)),founder=44,B=Blog)

x[,2:ncol(x)]<-x[,2:ncol(x)]/rowSums(x[,2:ncol(x)]) ## rescale to frequencies
gix <- x ## take the output to make a gif


##organise data for plotting
x <- x[,colnames(x)%in%c("time","(22)","(44)","(66)","(88)","(110)")]

y <- reshape2::melt(x,id.vars="time")
sinus_plot_data <- list(Bfunc_data = df_demo, c_crit_data = df_log,ode_data=y,ode_pars=ode_pars )
saveRDS(sinus_plot_data,"data/sinus_plot_data.Rds")


```
Plot the data

```{r}

hetero_plots <- function(plot_data,theta_index){
  pa <- ggplot(plot_data$Bfunc_data,aes(x=ploidy,y=B,color=theta,group=theta))+
    geom_line()+
    scale_x_continuous("copy number")+
    scale_y_log10("Mis-segr. rate (B)")+
    theme_classic(base_size=16)+
    scale_color_viridis_c(bquote(theta[.(theta_index)]))
  
  pb <- ggplot(plot_data$c_crit_data,aes(x=mu,y=beta))+
    geom_line()+
    geom_point(data=plot_data$ode_pars,size=2)+
    scale_x_continuous(expression(mu/lambda),
                       guide=guide_axis(check.overlap = TRUE))+
    scale_y_continuous(bquote(theta[.(theta_index)]))+
    theme_classic(base_size=16)
  
  
  pd <- ggplot(plot_data$c_crit_data,aes(x=mu,y=pop_B))+
    geom_line(aes(linetype="pop.\navg"))+
    geom_line(aes(y=minB,linetype="pop.\nminimum"))+
    scale_x_continuous(expression(mu/lambda),
                       guide=guide_axis(check.overlap = TRUE))+
    scale_y_log10(expression(beta))+
    theme_classic(base_size=16)+
    scale_linetype_discrete("")+
    theme(legend.position="top")

  
  
  pc <- ggplot(plot_data$ode_data,aes(x=time,y=value,color=variable))+
    geom_line()+
    scale_color_discrete("copy\nnumber")+
    theme_classic(base_size=16)
  
  if(theta_index==1){
    pc <- pc+scale_x_log10("days")+
      scale_y_continuous("relative frequency")
  }
  if(theta_index==2){
    pc <- pc+scale_x_log10("days")+#,limits=c(1,max(plot_data$ode_data$time)))+
    scale_y_log10("relative frequency",
                  limits=c(1e-6,1),
                  breaks=c(1,0.01,0.001,0.00001))
  }
  
  
  
  list(pa,pb,pc,pd)
  
}

x2 <- hetero_plots(readRDS("data/sinus_plot_data.Rds"),theta_index=2)
x1 <- hetero_plots(readRDS("data/lin_plot_data.Rds"),theta_index=1)


f6 <- egg::ggarrange(x1[[1]],x2[[1]],x1[[2]],x2[[2]],
                     x1[[3]],x2[[3]],x1[[4]],x2[[4]],
                     labels=c("A","B","C","D","E","F","G","H"),
                     ncol=2)
ggsave("figures/f6.png",plot=f6,width=9,height=12,unit="in")

```
make a gif of the ODE evolution for the 22<i<110 case
```{r}

z <- gix
colnames(z) <- c("time",2:ncol(z)+21)
z <- reshape2::melt(z,id.vars="time")

library(gganimate)

p <- ggplot(z,aes(x=as.numeric(variable)+21,y=value))+
  geom_point()+
  scale_y_log10("steady state ",limits=c(0.000001,1))+
  transition_time(log(time+0.00001))

anim_save("figures/ode_gid.gif",plot=p)

```