---
title: "CritCurvesOverlayed"
author: "Richard J Beck"
date: "6/8/2022"
output: html_document
---
This file generates figure showing the critical curves transformed to population average death and mis-segregation rates. Also the time to reach steady state
```{r setup}
knitr::opts_knit$set(root.dir = "PATH/TO/ROOT/DIR")
```

Source required functions and load libraries:
```{r,message=FALSE}
library(deSolve)
library(pbapply)
library(R.utils)
library(ggplot2)
library(ggpubr)
library(stringr)
library(deSolve)
library(egg)
source("Rscripts/functions.R")
source("Rscripts/plot_facets.R")
```

Show the heterogeneous functions we used:

```{r}

Bvar <- function(beta,i){
  b <- -0.09*i + beta*sin(45.59-12.25*log(i))
  pmin(1,exp(b)/100)
}

Dvar <- function(M,i) 1-(1-M)*(cos(2*pi*i/22)+1)/2

mu <- (1:10)/10
beta <- 2.66*seq(0.7,1.7,0.1)

cn <- 22:110

df1 <- do.call(rbind,lapply(mu, function(mi){
  data.frame(cn=cn,value=Dvar(mi,cn),id="mu",theta=mi)
}))
df2 <- do.call(rbind,lapply(beta, function(bi){
  data.frame(cn=cn,value=Bvar(bi,cn),id="beta",theta=bi)
}))


p2 <- ggplot(df1,aes(x=cn,y=value,color=theta,group=theta))+
  geom_line()+
  scale_x_continuous("copy number")+
  scale_y_continuous(expression(mu))+
  scale_color_viridis_c(expression(Mu))+
  theme_classic(base_size=12)
p2

p1 <- ggplot(df2,aes(x=cn,y=value,color=theta,group=theta))+
  geom_line()+
  scale_x_continuous("copy number")+
  scale_y_log10(expression(beta))+
  scale_color_viridis_c(expression(Beta))+
  theme_classic(base_size=12)
p1

p_funcs <- cowplot::plot_grid(p1,p2,labels = c("A","B"),ncol=2)
#ggsave("figures/hetero_curves.png",plot=p,width=7,height=3,units = "in")

```

generate critical curves using homogeneous and log functions:

```{r}
## function definitions (heterogeneous death and misseg)
Bvar <- function(beta,i){
  b <- -0.09*i + beta*sin(45.59-12.25*log(i))
  min(1,exp(b)/100)
}
#Dvar <- function(M,i) 1/(-0.002*i + M*sin(0.21 + 0.24 *log(i)))
Dvar <- function(M,i) 1-(1-M)*(cos(2*pi*i/22)+1)/2
maxchrom <- 110
minchrom <- 22

## set correct ranges for the B parameter
beta_hom <- seq(0.01,0.99,0.01) 
beta_lin <- 2.66*seq(0.7,1.7,0.1)

# critical curve for homogeneous mis-seg, heterogeneous death
#xhv --> x[homogenous][variable]
xhv <- do.call(rbind,pblapply(c(0.01,seq(0.1,0.9,0.2),0.99), function(beta) find_c_crit(beta,maxchrom=110,minchrom=22,D=Dvar)))

## same notation idea as above
xvv <- do.call(rbind,pblapply(beta_lin, function(beta) find_c_crit(beta,maxchrom=110,minchrom=22,D=Dvar,B=Bvar)))

##...
xhh <- pbsapply(beta_hom, function(beta) fast_c_crit(beta,maxchrom=110,minchrom=22))
xvh <- pbsapply(beta_lin, function(beta) fast_c_crit(beta,maxchrom=110,minchrom=22,B=Bvar))

##compute the population average rates
dfhv <- data.frame(beta=c(0.01,seq(0.1,0.9,0.2),0.99),mu=xhv$mu,Btype="homo",Dtype="hetero")
dfhv$popB <- dfhv$beta ##homogeneous function = constant rate
## heterogeneous death: rate needs computed
dfhv$popD <- sapply(1:nrow(dfhv), function(i) {
  pop_ss_D(beta=dfhv$beta[i],mu=dfhv$mu[i],maxchrom=maxchrom,
           minchrom=minchrom,D=Dvar)})
## etc as above...                    
dfvv <- data.frame(beta=beta_lin,mu=xvv$mu,Btype="hetero",Dtype="hetero") 
dfvv$popB <- sapply(1:nrow(dfvv), function(i) {
  pop_ss_B(beta=dfvv$beta[i],mu=dfvv$mu[i],maxchrom=maxchrom,
           minchrom=minchrom,B=Bvar,D=Dvar)})

dfvv$popD <- sapply(1:nrow(dfvv), function(i) {
  pop_ss_D(beta=dfvv$beta[i],mu=dfvv$mu[i],maxchrom=maxchrom,
           minchrom=minchrom,B=Bvar,D=Dvar)})
                     
dfhh <- data.frame(beta=beta_hom,mu=xhh,Btype="homo",Dtype="homo")
dfhh$popB <- dfhh$beta
dfhh$popD <- dfhh$mu
dfvh <- data.frame(beta=beta_lin,mu=xvh,Btype="hetero",Dtype="homo")
dfvh$popB <- pbsapply(1:nrow(dfvh), function(i) {
  pop_ss_B(beta=dfvh$beta[i],mu=dfvh$mu[i],maxchrom=maxchrom,
           minchrom=minchrom,B=Bvar)})
dfvh$popD <- dfvh$mu

##compile the data for plotting
df <- rbind(dfvv,dfvh,dfhh,dfhv)
saveRDS(df,"data/CritCurvesOverlayed.Rds")

```
Plot the data

```{r}
df <- readRDS("data/CritCurvesOverlayed.Rds")


df$Btype <- sapply(df$Btype, function(x){
  eq <- c("beta==Beta",
          "beta==-0*.*09*i+Beta*sin(45.6-12.3*log~i)")[x==c("homo","hetero")]
 #         "Beta==exp(beta-1.352857*i)/100")[x==c("homo","hetero")]
})
df$Dtype <- sapply(df$Dtype, function(x){
  eq <- c("mu==Mu",
          "mu==1-(1-Mu)*(cos(2*pi*i/22)+1)/2")[x==c("homo","hetero")]
})

types <- expand.grid(Btype=unique(df$Btype),
                     Dtype=unique(df$Dtype))

## a lot of hassle to get the plot labels - https://stackoverflow.com/a/52708749/8145482
p1 <- ggplot(df,aes(x=Re(popD),y=Re(popB)))+
  facet_wrap(c("Btype","Dtype"),scales="free",labeller=label_parsed)+
  geom_line()+
  scale_x_continuous(expression(pop.~avg.~death~rate~(mu[pop])),
                     guide=guide_axis(check.overlap = TRUE))+
  scale_y_continuous(expression(pop.~avg.~missegr.~rate~(beta[pop])))+
  theme_classic(base_size=12)+
  theme(strip.background = element_blank(),
        strip.text = element_blank())


p2 <- ggplot(df,aes(x=Re(popD),y=Re(popB)))+
  facet_grid(c("Btype","Dtype"),scales="free",labeller=label_parsed)+
  geom_line()+
  scale_x_continuous(expression(pop.~avg.~death~rate~(mu[pop])),
                     guide=guide_axis(check.overlap = TRUE))+
  scale_y_continuous(expression(pop.~avg.~missegr.~rate~(beta[pop])))+
  theme_classic(base_size=12)+
    theme(strip.background = element_blank(),
        strip.text.y = element_text(size=8)
          )

p_ccrit <- plot_facets(p1,p2)
grid.newpage()
grid.draw(p_ccrit)

#ggsave("figures/CritCurvesOverlayed.png",plot=p_ccrit,width=7,height=5,unit="in")



```
Run the ODE model with parameter combinations along the critical curves.
Quantify the distance between the ODE output at each time point, and the steady state
```{r}
## define functions
Bvar <- function(beta,i){
  b <- -0.09*i + beta*sin(45.59-12.25*log(i))
  min(1,exp(b)/100)
}
Dvar <- function(M,i) 1-(1-M)*(cos(2*pi*i/22)+1)/2

Bhom <- function(beta,i) return(beta)
Dhom <- function(M,i) return(M)

## function to get difference between ODE at time t (over a wide range of t) and steady state
ttss <- function(beta,mu,B,D,minchrom,maxchrom){
  ## first figure out what the steady state is
  A <- get_A(maxchrom=maxchrom,minchrom=minchrom,beta=beta,mu=mu,B=B,D=D)
  ss <- eigen(t(A))
  ss <- Re(ss$vectors[,Re(ss$values)==max(Re(ss$values))])
  ss <- ss/sum(ss)
  
  ## run the model over a large range of times
  times <- seq(1,99.9,0.1)
  times <- c(times,times*100,times*100^2,times*100^3,times*100^4)
  
  x <- run_ode(beta=beta,mu=mu,maxchrom=maxchrom,founder=44,minchrom=minchrom,
                B=B,D=D,times = times)[,-1]
  x <- x/rowSums(x)
  ## intergrate the difference between ODE state and steady state over time
  fin <- apply(x,1, function(i){
    sum(abs(i-ss))
  })
  
  data.frame(times=times,dist=fin)
  
  
}
## load  parameter sets on the critical curves
df <- readRDS("data/CritCurvesOverlayed.Rds")
df <- df[df$Btype=="hetero"|df$Dtype=="hetero"|df$beta%in%c(0.01,seq(0.05,0.95,0.05),0.99),]

## run ODE for each parameter set and compute distances to steady state
ttss <- do.call(rbind,pblapply(1:nrow(df), function(i){
  B <- c(Bvar,Bhom)[df$Btype[i]==c("hetero","homo")][[1]]
  D <- c(Dvar,Dhom)[df$Dtype[i]==c("hetero","homo")][[1]]
  xi <- ttss(df$beta[i],df$mu[i],B,D,22,110)
  xi$Btype <-df$Btype[i]
  xi$Dtype <- df$Dtype[i]
  xi$beta <- df$beta[i]
  xi$mu <- df$mu[i]
  xi$popB <- df$popB[i]
  xi$popD <- df$popB[i]
  xi
}))

saveRDS(ttss,"data/ttss.Rds")


```

Filter results from prior chunk and plot output
```{r}
df <- readRDS("data/ttss.Rds")

df <- df[df$dist<0.001,] ## keep only ODE output that is "sufficiently" close to steady state (note that the distance generally goes from 2->1, so a dist of 0.001 is close indeed)

## for each of the four conditions, and for each point on the critical curve identified by the parameter B (called beta in code), return the first timepoint where the modely was "sufficiently" close to steady state
df <- aggregate(list(time=df$times),by=list(beta=df$popB,
                                            Dtype=df$Dtype,
                                            Btype=df$Btype),min)


## make nice labels for plotting
df$Btype <- sapply(df$Btype, function(x){
  eq <- c("beta==Beta",
          "beta==-0*.*09*i+Beta*sin(45.6-12.3*log~i)")[x==c("homo","hetero")]
 #         "Beta==exp(beta-1.352857*i)/100")[x==c("homo","hetero")]
})
df$Dtype <- sapply(df$Dtype, function(x){
  eq <- c("mu==Mu",
          "mu==1-(1-M)*(cos(2*pi*i/22)+1)/2")[x==c("homo","hetero")]
})



p1 <- ggplot(df,aes(x=beta,y=time))+
  facet_wrap(c("Btype","Dtype"),scales="free",labeller=label_parsed)+
  geom_line()+
  scale_x_log10(expression(pop.~avg.~missegr.~rate~(beta[pop])))+
  scale_y_log10("time to steady state (days)")+
  theme_classic(base_size=12)+
  theme(strip.background = element_blank(),
        strip.text = element_blank())


p2 <- ggplot(df,aes(x=beta,y=time))+
  facet_grid(c("Btype","Dtype"),scales="free",labeller=label_parsed)+
  geom_line()+
  scale_x_log10(expression(pop.~avg.~missegr.~rate~(beta[pop])))+
  scale_y_log10("time to steady state (days)")+
  theme_classic(base_size=12)+
    theme(strip.background = element_blank(),
        strip.text.y = element_text(size=8)
          )

p_ttss <- plot_facets(p1,p2)
grid.newpage()
grid.draw(p_ttss)

#ggsave("figures/ttss.png",plot=p_ttss,width=7,height=5,unit="in")
```
Compile all the subpanels and plot
```{r}

cco <- cowplot::plot_grid(p_funcs,p_ccrit,p_ttss,labels = c("","C","D"),rel_heights = c(3,6,6),ncol=1)
ggsave("figures/CritCurvesOverlayed.png",plot=cco,width=7,height=10,unit="in",bg = "white")
```

