---
title: "Critical Curves"
author: "Richard J Beck"
date: "6/8/2022"
output: html_document
---

This file produces figures for critical curves where the mis-segregation rate is homogeneous. Some validation of the critical curves is also performed.

Set working directory

```{r setup}
knitr::opts_knit$set(root.dir = "PATH/TO/ROOT/DIR")
```

Source required functions and load libraries:
```{r,message=FALSE}
library(deSolve)
library(pbapply)
library(R.utils)
library(ggplot2)
library(ggpubr)
library(stringr)
library(deSolve)
library(egg)
source("Rscripts/functions.R")
```

Generate critical curves:
```{r}

beta_vals <- seq(0.05,0.95,0.05)

df5 <- do.call(rbind,pblapply(beta_vals,fast_c_crit,maxchrom=5,df=TRUE))
df10 <- do.call(rbind,pblapply(beta_vals,fast_c_crit,maxchrom=10,df=TRUE))
df510 <- do.call(rbind,pblapply(beta_vals,fast_c_crit,maxchrom=c(5,10),df=TRUE))
df_agg <- do.call(rbind,pblapply(beta_vals,fast_c_crit,maxchrom=110,minchrom=22,df=TRUE))
df <- rbind(df5,df10,df510,df_agg)

saveRDS(df,"data/critical_curves.Rds")

```

Perform numerical validation of critical curves
```{r} 

test_growth <- function(beta,mu,maxchrom){
  minchrom <- 1
  founder <- 2
  mc<- as.numeric(unlist(strsplit(maxchrom,",")))
  if(110%in%mc){
    minchrom <- 22
    founder <- 44
  }
  growth <- check_growth(beta,mu,mc,founder,minchrom)
  data.frame(beta=beta,mu=mu,growth=growth,maxchrom=maxchrom)
}

pars <- expand.grid(mu=seq(0.05,0.95,0.1),beta=seq(0.05,0.95,.1))


x5 <- do.call(rbind,pblapply(1:nrow(pars), function(i){
  test_growth(pars$beta[i],pars$mu[i],"5")
}))
x10 <- do.call(rbind,pblapply(1:nrow(pars), function(i){
  test_growth(pars$beta[i],pars$mu[i],"10")
}))
x510 <- do.call(rbind,pblapply(1:nrow(pars), function(i){
  test_growth(pars$beta[i],pars$mu[i],"5,10")
}))

xagg <- do.call(rbind,pblapply(1:nrow(pars), function(i){
  test_growth(pars$beta[i],pars$mu[i],"110")
}))

x <- rbind(x5,x10,x10,x510,xagg)
saveRDS(x,"data/critical_curves_numerical.Rds")
```
Plot critical curves overlayed with numerical validation
```{r}
df <- readRDS("data/critical_curves.Rds")
x <- readRDS("data/critical_curves_numerical.Rds")
relablr <- function(mc) {
  c('5'=" 1 <= i[x] ~ phantom() <= 5",'10'="1 <= i[x] ~ phantom() <= 10",
    '5,10'="1 <= i[x1] ~ phantom() <= 5~and~1 <= i[x2] ~ phantom() <= 10",
    '110'="22 <= ploidy ~ phantom() <= 110")[mc]
}
x$id <- sapply(x$maxchrom,relablr)
df$id <- sapply(df$maxchrom,relablr)
p <- ggplot(df,aes(x=mu,y=beta))+
  facet_wrap(c("id"),labeller=label_parsed)+
  geom_point(data=x,aes(shape=growth))+
  geom_path()+
  theme_classic(base_size=12)+
  scale_x_continuous(expression(death~rate~(mu)))+
  scale_y_continuous(expression(mis-segr.~rate~(beta)))+
  scale_shape_discrete("population\nexpands")
p
ggsave("figures/ccrit_numerical.png",plot=p,width=7,height=5,unit="in")
```
Using the fast method for finding critical curves, generate critical curves for multiple chromosomes
```{r}

beta <- c(seq(0.01,0.99,0.01)/100,seq(0.01,0.99,0.01))
maxchrom <- c(lapply(1:22, function(i) rep(3,i)),
              lapply(1:22, function(i) rep(5,i)),
              lapply(1:22, function(i) rep(8,i)))



df <- do.call(rbind,pblapply(maxchrom, function(mc){
  mu <- sapply(beta,very_fast_c_crit,maxchrom=mc[1],n=length(mc))
  data.frame(beta=beta,mu=mu,maxchrom=mc[1],nchrom=length(mc))
  
}))

saveRDS(df,"data/multi_ccrit.Rds")

```

Plot multiple chromosome results:
```{r}

df <- readRDS("data/multi_ccrit.Rds")

lablr <- function(x) paste0("1 <= i[x] ~ phantom() <=",x)
df$label <- sapply(df$maxchrom,lablr)

fig_subset <- function(mc){
  p <- ggplot(subset(df,maxchrom==mc),aes(x=Re(mu),y=beta,color=nchrom,group=nchrom))+
  facet_grid(cols=vars(label),labeller=label_parsed)+
  geom_path()+
  scale_x_continuous(expression(mu/lambda),limits=c(0,1))+
  scale_y_continuous(expression(beta))+
  scale_color_viridis_c(expression(k))+
  theme_classic(base_size=12)
  if(mc<8) p <- p+theme(legend.position="none")
p
}

f <- lapply(c(3,5,8),fig_subset)

cowplot::plot_grid(f[[1]],f[[2]],f[[3]],ncol=3,rel_widths = c(2,2,2.8),labels=c("A","B","C"))

ggsave("figures/multichrom_ccrit.png",width=7,height=3,unit="in")
```

Test that the three different methods for validating critical curves produce identical results:
```{r}
beta <- seq(0.05,0.95,0.1)
maxchrom <- lapply(1:3, function(i) rep(3,i))

df3 <- do.call(rbind,pblapply(beta,find_c_crit,maxchrom=3))



df <- do.call(rbind,lapply(maxchrom, function(mc){
  x <- do.call(rbind,pblapply(beta,find_c_crit,maxchrom=mc))
  x$mu_fast <- pbsapply(beta,fast_c_crit,maxchrom=mc)
  x$mu_vfast <- pbsapply(beta,very_fast_c_crit,maxchrom=mc[1],n=length(mc))
  return(x)
}))
df <- df[df$v<0.00001,]
df <- df[,-1]
df <- reshape2::melt(df,id.vars=c("beta","maxchrom"))
saveRDS(df,"data/ccrit_methods_test.Rds")

```

Plot validation results
```{r}

df <- readRDS("data/ccrit_methods_test.Rds")

p <- ggplot(df,aes(x=value,y=beta,color=maxchrom))+
  facet_grid(cols=vars(variable))+
  geom_path(size=2)+
  scale_x_continuous(expression(mu/lambda))+
  scale_y_continuous(expression(beta))
p
```


